---
format: gfm
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# imaginarycss: Tools for Studying Imaginary Cognitive Social Structure

<!-- badges: start -->
```{=markdown}
[![CRAN status](https://www.r-pkg.org/badges/version/imaginarycss)](https://CRAN.R-project.org/package=imaginarycss)
[![R-CMD-check](https://github.com/gvegayon/imaginarycss/actions/workflows/r-cmd-check.yml/badge.svg)](https://github.com/gvegayon/imaginarycss/actions/workflows/r-cmd-check.yml)
[![CRANlogs downloads](https://cranlogs.r-pkg.org/badges/grand-total/imaginarycss)](https://cran.r-project.org/package=imaginarycss)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://github.com/gvegayon/imaginarycss/blob/master/LICENSE.md)
[![codecov](https://codecov.io/gh/gvegayon/imaginarycss/graph/badge.svg?token=ZB8FVLI7GN)](https://app.codecov.io/gh/gvegayon/imaginarycss)
[![status](https://tinyverse.netlify.app/badge//imaginarycss)](https://CRAN.R-project.org/package=imaginarycss)
```
<!-- badges: end -->

The `imaginarycss` package provides functions to measure and test imaginary cognitive social structure (CSS) motifs, which are patterns of perceived relationships among individuals in a social network. The package includes tools for calculating motif frequencies, comparing observed motifs to expected distributions, and visualizing motif structures. It implements the methods described in Tanaka and Vega Yon (2023) <DOI:10.1016/j.socnet.2023.11.005>.

When using this package, please cite the above article:

```{r}
#| label: citation
#| comment: ''
citation(package = "imaginarycss")
```

## Installation

You can install the development version of `imaginarycss` from [GitHub](https://github.com/) with:

``` r
devtools::install_github("gvegayon/imaginarycss")
```

Or from [R-universe](https://gvegayon.r-universe.dev/){target="_blank"} (recommended for the latest development version):

```r
install.packages(
  'imaginarycss',
  repos = c(
    'https://gvegayon.r-universe.dev',
    'https://cloud.r-project.org'
  )
)
```

Or from CRAN

```r
install.packages("imaginarycss")
```


## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(imaginarycss)

source_ <- c(1, 2, 3, 1) 
target_ <- c(2, 1, 4, 4)

source_ <- c(source_, source_[-1] + 4)
target_ <- c(target_, target_[-1] + 4)

adjmat <- matrix(0L, nrow = 8, ncol = 8)
adjmat[cbind(source_, target_)] <- 1L

graph <- new_barry_graph(adjmat, n = 4)
graph

# These are two attributes that are part of the barry_graph object
attr(graph, "endpoints")
attr(graph, "netsize")

count_recip_errors(graph)
```

## Code of Conduct
  
Please note that the imaginarycss project is released with a [Contributor Code of Conduct](https://gvegayon.github.io/imaginary-structures/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
