---
title: "Network Perception Analysis"
author: "Sima Najafzadehkhoei"
date: today
vignette: >
  %\VignetteIndexEntry{Network Perception Analysis}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# Imaginary Network Motifs: Testing Systematic Perception Errors

People do not perceive social networks randomly---their errors follow
systematic patterns (Tanaka & Vega Yon, 2023). This vignette demonstrates
how to use `imaginarycss` to (1) compute observed imaginary-census motif
counts, (2) generate a null distribution that preserves each perceiver's
tie-level accuracy, and (3) test which motif patterns are significantly
elevated or reduced relative to chance.

## Setup

```{r setup, message=FALSE}
library(imaginarycss)

# Load Krackhardt data
data(krack_advice)
data(krack_friendship)
data(Krack_friendship_perceptions)
data(Krack_advice_perceptions)

# Convert edge-list data frames to adjacency matrices
df_to_adjmat <- function(df) {
  n <- max(c(df$from, df$to))
  m <- matrix(0L, nrow = n, ncol = n)
  m[cbind(df$from, df$to)] <- df$value
  m
}

advice_matrix     <- df_to_adjmat(krack_advice)
friendship_matrix <- df_to_adjmat(krack_friendship)
```

## Data Overview

```{r data-overview}
n <- nrow(advice_matrix)
cat("Dataset: 21 managers in a high-tech company\n")
cat("Friendship -- Edges:", sum(friendship_matrix),
    "| Density:", round(sum(friendship_matrix) / (n * (n - 1)), 3), "\n")
cat("Advice     -- Edges:", sum(advice_matrix),
    "| Density:", round(sum(advice_matrix) / (n * (n - 1)), 3), "\n")
```

## Creating the CSS Graphs

Each graph bundles the true network (first element) with every individual's
perception.

```{r graphs}
friendship_graph <- new_barry_graph(
  c(list(friendship_matrix), krack_friendship_perceptions)
)
advice_graph <- new_barry_graph(
  c(list(advice_matrix), krack_advice_perceptions)
)
```

## Observed Motif Counts

The imaginary census classifies every perceiver--dyad pair into one of ten
categories. We aggregate across perceivers to get total counts per motif.

```{r observed}
friendship_observed <- count_imaginary_census(friendship_graph)
advice_observed     <- count_imaginary_census(advice_graph)

# Helper: aggregate census by motif type
aggregate_motifs <- function(census) {
  motif <- gsub("\\([0-9]+\\)\\s*", "", census$name)
  tapply(census$value, motif, sum)
}

cat("Friendship motifs (top 6):\n")
fr_agg <- sort(aggregate_motifs(friendship_observed), decreasing = TRUE)
print(head(fr_agg))

cat("\nAdvice motifs (top 6):\n")
ad_agg <- sort(aggregate_motifs(advice_observed), decreasing = TRUE)
print(head(ad_agg))
```

## Null-Model Testing

We generate 100 null CSS samples for each network. Each sample preserves
every perceiver's four accuracy rates (true-positive/true-negative for
ego/alter dyads) but randomises which specific dyads are misperceived.

```{r null-testing}
set.seed(331)

null_motifs <- function(graph, n_sim = 100) {
  replicate(n_sim, {
    aggregate_motifs(
      count_imaginary_census(
        new_barry_graph(sample_css_network(graph))
      )
    )
  })
}

friendship_null <- null_motifs(friendship_graph)
advice_null     <- null_motifs(advice_graph)
```

## Statistical Significance

For each motif we compute a z-score and a two-sided empirical p-value
comparing the observed count to the null distribution.

```{r significance}
test_motifs <- function(observed_agg, null_matrix) {
  motifs   <- names(observed_agg)
  null_mat <- null_matrix[motifs, , drop = FALSE]
  data.frame(
    motif     = motifs,
    observed  = as.integer(observed_agg),
    null_mean = round(rowMeans(null_mat), 1),
    null_sd   = round(apply(null_mat, 1, sd), 2),
    z_score   = round(
      (observed_agg - rowMeans(null_mat)) / apply(null_mat, 1, sd), 2
    ),
    p_value   = round(2 * pmin(
      rowMeans(null_mat <= observed_agg),
      rowMeans(null_mat >= observed_agg)
    ), 4),
    row.names = NULL,
    stringsAsFactors = FALSE
  )
}

friendship_results <- test_motifs(fr_agg, friendship_null)
advice_results     <- test_motifs(ad_agg, advice_null)

friendship_results$significant <- abs(friendship_results$z_score) > 1.96 &
  friendship_results$p_value < 0.05
advice_results$significant     <- abs(advice_results$z_score) > 1.96 &
  advice_results$p_value < 0.05
```

## Results

```{r results, fig.height=7, fig.width=8}
show_significant <- function(results, label) {
  sig <- results[results$significant, ]
  if (nrow(sig) > 0) {
    cat(label, "-- significant motifs:\n")
    print(sig[order(-abs(sig$z_score)),
              c("motif", "observed", "z_score", "p_value")])
  } else {
    cat(label, "-- no significant deviations\n")
  }
  cat("\n")
}

show_significant(friendship_results, "Friendship")
show_significant(advice_results,     "Advice")

# Visualise z-scores for both networks
plot_zscores <- function(results, main) {
  results <- results[order(results$z_score), ]
  cols <- ifelse(results$significant, "steelblue", "grey80")
  par(mar = c(4, 12, 3, 1))
  bp <- barplot(
    results$z_score,
    horiz = TRUE, las = 1,
    names.arg = results$motif,
    col = cols, border = NA,
    xlab = "Z-Score", main = main
  )
  abline(v = c(-1.96, 1.96), lty = 2, col = "red")
  abline(v = 0, col = "grey50")
}

plot_zscores(friendship_results, "Friendship: Motif Z-Scores vs Null")
plot_zscores(advice_results,     "Advice: Motif Z-Scores vs Null")
```

## Tie-Level Accuracy

```{r accuracy}
friendship_acc <- tie_level_accuracy(friendship_graph)
advice_acc     <- tie_level_accuracy(advice_graph)

cat("Mean individual accuracy:\n")
cat(sprintf("  Friendship -- TP: %.3f  TN: %.3f\n",
  mean(friendship_acc$p_1_ego, na.rm = TRUE),
  mean(friendship_acc$p_0_ego, na.rm = TRUE)))
cat(sprintf("  Advice     -- TP: %.3f  TN: %.3f\n",
  mean(advice_acc$p_1_ego, na.rm = TRUE),
  mean(advice_acc$p_0_ego, na.rm = TRUE)))
```

## Conclusion

By comparing observed motif counts against null distributions that preserve
individual-level accuracy, the imaginary census reveals *which* perceptual
errors are systematic rather than random. Elevated motifs indicate cognitive
schemas (e.g., balance, reciprocity) that distort network perception beyond
what individual accuracy rates alone would predict.
