---
title: "Network Perception Analysis"
author: "Sima Najafzadehkhoei"
date: today
vignette: >
  %\VignetteIndexEntry{Network Perception Analysis}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

# Imaginary Network Motifs: Analyzing Systematic Perception Errors

Understanding how people perceive social networks reveals systematic biases in human cognition. This vignette demonstrates the imaginary network motifs framework using the classic Krackhardt dataset to identify when perceived network patterns deviate significantly from chance expectations.

## Setup

```{r setup, message=FALSE}
library(imaginarycss)
library(ggplot2)
library(dplyr)
library(reshape2)

# Load Krackhardt high-tech managers data
data(krack_advice)
data(krack_friendship) 
data(Krack_friendship_perceptions)
data(Krack_advice_perceptions)

# Convert to adjacency matrices
advice_matrix <- acast(krack_advice, from ~ to, value.var = "value", fill = 0)
friendship_matrix <- acast(krack_friendship, from ~ to, value.var = "value", fill = 0)
```

## Data Overview

```{r data-overview}
cat("Dataset: 21 managers in a high-tech company\n")
cat("Friendship network - Edges:", sum(friendship_matrix), 
    "| Density:", round(sum(friendship_matrix)/(21*20), 3), "\n")
cat("Advice network - Edges:", sum(advice_matrix), 
    "| Density:", round(sum(advice_matrix)/(21*20), 3), "\n")
cat("Perception networks per type:", length(krack_friendship_perceptions))
```

## Creating Network Graphs

```{r graphs}
# Combine true networks with individual perceptions
friendship_graph <- new_barry_graph(c(list(friendship_matrix), krack_friendship_perceptions))
advice_graph <- new_barry_graph(c(list(advice_matrix), krack_advice_perceptions))
```

## Observed Motif Patterns

```{r observed-analysis}
# Count imaginary census motifs
friendship_observed <- count_imaginary_census(friendship_graph)
advice_observed <- count_imaginary_census(advice_graph)

# Clean and aggregate motif data
process_motifs <- function(census_data) {
  census_data %>%
    mutate(motif_type = gsub("\\([0-9]+\\)$", "", name)) %>%
    group_by(motif_type) %>%
    summarise(
      total = sum(value),
      mean_per_person = mean(value),
      variation = var(value),
      .groups = 'drop'
    ) %>%
    arrange(desc(total))
}

friendship_summary <- process_motifs(friendship_observed)
advice_summary <- process_motifs(advice_observed)

# Show top patterns
cat("Top friendship motifs:\n")
print(head(friendship_summary[, 1:3], 6))
cat("\nTop advice motifs:\n")
print(head(advice_summary[, 1:3], 6))
```

## Distribution Visualization

```{r motif-distributions, fig.height=8, fig.width=12}
# Create distribution plot
create_distribution_plot <- function(census_data, network_name) {
  census_clean <- census_data %>%
    mutate(motif_type = gsub("\\([0-9]+\\)$", "", name))
  
  ggplot(census_clean, aes(x = value)) +
    geom_histogram(bins = 8, fill = "steelblue", alpha = 0.7, color = "white") +
    facet_wrap(~motif_type, scales = "free", labeller = label_wrap_gen(width = 20)) +
    labs(title = paste("Distribution of Motif Types -", network_name),
         subtitle = "Individual-level counts across 21 managers",
         x = "Count", y = "Frequency") +
    theme_minimal() +
    theme(
      strip.text = element_text(size = 9, face = "bold"),
      plot.title = element_text(size = 14, face = "bold"),
      axis.text = element_text(size = 8)
    )
}

# Plot distributions
friendship_dist_plot <- create_distribution_plot(friendship_observed, "Friendship")
print(friendship_dist_plot)
```

## Null Model Testing

```{r null-testing}
# Generate null distributions via CSS sampling
cat("Generating null models (100 samples each)...\n")

set.seed(331)
friendship_null <- replicate(100, {
  count_imaginary_census(new_barry_graph(sample_css_network(friendship_graph)))
}, simplify = FALSE)

advice_null <- replicate(100, {
  count_imaginary_census(new_barry_graph(sample_css_network(advice_graph)))
}, simplify = FALSE)
```

## Statistical Significance Analysis

```{r significance}
# Calculate Z-scores against null expectation
test_significance <- function(observed, null_samples) {
  # Aggregate observed by motif type
  obs_agg <- observed %>%
    mutate(motif = gsub("\\([0-9]+\\)$", "", name)) %>%
    group_by(motif) %>%
    summarise(observed = sum(value), .groups = 'drop')
  
  # Process null samples 
  null_matrix <- sapply(null_samples, function(x) {
    x %>% 
      mutate(motif = gsub("\\([0-9]+\\)$", "", name)) %>%
      group_by(motif) %>%
      summarise(null_val = sum(value), .groups = 'drop') %>%
      arrange(motif) %>%
      pull(null_val)
  })
  
  # Statistical testing
  obs_agg %>%
    mutate(
      null_mean = rowMeans(null_matrix),
      null_sd = apply(null_matrix, 1, sd),
      z_score = (observed - null_mean) / null_sd,
      p_value = 2 * pmin(
        rowMeans(null_matrix <= observed),
        rowMeans(null_matrix >= observed)
      ),
      significant = abs(z_score) > 1.96 & p_value < 0.05,
      direction = ifelse(z_score > 0, "Elevated", "Reduced")
    ) %>%
    arrange(desc(abs(z_score)))
}

# Run tests
friendship_results <- test_significance(friendship_observed, friendship_null)
advice_results <- test_significance(advice_observed, advice_null)
```

## Key Findings

```{r results, fig.height=10, fig.width=12}
# Display significant results
cat("=== SIGNIFICANT DEVIATIONS FROM CHANCE ===\n\n")

show_significant <- function(results, network_name) {
  sig_results <- results[results$significant, ]
  if(nrow(sig_results) > 0) {
    cat(network_name, "Network:\n")
    print(sig_results[, c("motif", "observed", "z_score", "p_value", "direction")])
    cat("\n")
  } else {
    cat(network_name, "Network: No significant deviations\n\n")
  }
}

show_significant(friendship_results, "Friendship")
show_significant(advice_results, "Advice")

# Visualization
top_n <- 8
combined <- rbind(
  data.frame(friendship_results[1:top_n, ], Network = "Friendship"),
  data.frame(advice_results[1:top_n, ], Network = "Advice")
)

significance_plot <- ggplot(combined, aes(x = reorder(motif, z_score), y = z_score)) +
  geom_col(aes(fill = significant), alpha = 0.8, width = 0.7) +
  geom_hline(yintercept = c(-1.96, 1.96), linetype = "dashed", 
             color = "red", alpha = 0.8, size = 0.8) +
  geom_hline(yintercept = 0, color = "black", alpha = 0.4) +
  facet_wrap(~Network, scales = "free", ncol = 1) +
  coord_flip() +
  scale_fill_manual(
    values = c("FALSE" = "lightgray", "TRUE" = "steelblue"),
    name = "Significant\n(p < 0.05)"
  ) +
  labs(
    title = "Motif Significance Analysis",
    subtitle = "Z-scores vs null model | Red lines = significance threshold (±1.96)",
    x = "Motif Type",
    y = "Z-Score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text.y = element_text(size = 10),
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "bottom"
  )

print(significance_plot)
```

## Interpretation

```{r interpretation}
# Summary statistics
friend_sig_count <- sum(friendship_results$significant)
advice_sig_count <- sum(advice_results$significant)
friend_elevated <- sum(friendship_results$significant & friendship_results$z_score > 0)
advice_elevated <- sum(advice_results$significant & advice_results$z_score > 0)

cat("=== SUMMARY ===\n")
cat("Friendship network:", friend_sig_count, "significant motifs (", 
    friend_elevated, "elevated,", friend_sig_count - friend_elevated, "reduced)\n")
cat("Advice network:", advice_sig_count, "significant motifs (",
    advice_elevated, "elevated,", advice_sig_count - advice_elevated, "reduced)\n")

# Effect sizes
cat("\nEffect sizes (mean |Z-score|):\n")
cat("Friendship:", round(mean(abs(friendship_results$z_score)), 2), "\n")
cat("Advice:", round(mean(abs(advice_results$z_score)), 2), "\n")

# Theoretical implications
cat("\n=== THEORETICAL IMPLICATIONS ===\n")
if(friend_sig_count > advice_sig_count) {
  cat("• Friendship networks show stronger systematic biases\n")
  cat("• Consistent with balance schema and reciprocity assumptions\n")
} else {
  cat("• Advice networks show stronger systematic biases\n")
  cat("• May reflect hierarchy and expertise perception patterns\n")
}

cat("• Individual perception varies significantly from random baselines\n")
cat("• Cognitive schemas systematically distort network perception\n")
```

## Tie-Level Accuracy

```{r accuracy}
# Individual accuracy analysis
friendship_acc <- tie_level_accuracy(friendship_graph)
advice_acc <- tie_level_accuracy(advice_graph)

# Quick comparison
acc_summary <- data.frame(
  Network = c("Friendship", "Advice"),
  True_Positive_Rate = c(mean(friendship_acc$p_1_ego, na.rm=T), 
                        mean(advice_acc$p_1_ego, na.rm=T)),
  True_Negative_Rate = c(mean(friendship_acc$p_0_ego, na.rm=T),
                        mean(advice_acc$p_0_ego, na.rm=T))
)

cat("Individual-level accuracy:\n")
print(acc_summary)
```

## Conclusion

This analysis demonstrates how the imaginary network motifs framework reveals systematic biases in social network perception. By comparing observed patterns against null models generated via `sample_css_network()`, we identify which cognitive errors occur beyond chance expectations.

**Key methodological advances:**

 - **Null model rigor**: Random network comparison provides statistical baseline 
 - **Individual variation**: Framework captures person-level differences in perception accuracy\
- **Motif-level analysis**: Goes beyond simple accuracy to examine structural error patterns

 - **Systematic bias detection**: Identifies cognitive schemas that distort network perception

The results illuminate how human cognition systematically misperceives social structure, with implications for organizational behavior, social psychology, and network analysis methodology.
