---
title: "Real-World Example: Krackhardt Advice Network"
author: "Sima Najafzadehkhoei"
date: today
vignette: >
  %\VignetteIndexEntry{Real-world example of imaginarycss}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# Krackhardt Advice Network

This vignette demonstrates a complete `imaginarycss` workflow using the
classic Krackhardt (1987) dataset of 21 managers in a high-tech company.
We construct a CSS object from the advice network together with simulated
perceptual errors, then compute accuracy rates and the imaginary census.

```{r setup}
library(imaginarycss)
set.seed(42)
```

## Data and Perceptions

The package ships the advice network as a data frame (`krack_advice`).
We convert it to an adjacency matrix using base R and then generate
realistic perception errors driven by three cognitive biases:

- **Visibility** -- peripheral ties are harder to perceive (false
  negatives).
- **Status** -- high-indegree nodes attract spurious perceived ties
  (false positives).
- **Reciprocity** -- people tend to assume that observed ties are
  reciprocated.

```{r data}
data(krack_advice)

# Convert edge-list data frame to adjacency matrix
n_people <- max(c(krack_advice$from, krack_advice$to))
advice_matrix <- matrix(0L, nrow = n_people, ncol = n_people)
advice_matrix[cbind(krack_advice$from, krack_advice$to)] <- krack_advice$value

# Generate realistic perceptions with three bias mechanisms
create_realistic_perceptions <- function(true_net) {
  n <- nrow(true_net)
  out <- vector("list", n)
  colsums <- colSums(true_net)
  highdeg <- colsums > median(colsums)
  for (i in seq_len(n)) {
    P <- true_net
    for (j in seq_len(n)) for (k in seq_len(n)) if (i != j && i != k) {
      if (true_net[j, k] == 1 && runif(1) < 0.20) P[j, k] <- 0
      if (true_net[j, k] == 0 && highdeg[k] && runif(1) < 0.15) P[j, k] <- 1
      if (true_net[j, k] == 1 && true_net[k, j] == 0 && runif(1) < 0.10) P[k, j] <- 1
    }
    diag(P) <- 0
    out[[i]] <- P
  }
  out
}

perceived_networks <- create_realistic_perceptions(advice_matrix)
krack_graph <- new_barry_graph(c(list(advice_matrix), perceived_networks))
```

## Individual Accuracy

```{r accuracy, fig.height=5, fig.width=6}
acc <- tie_level_accuracy(krack_graph)

# Summary statistics
cat("Mean accuracy rates across", nrow(acc), "perceivers:\n")
cat("  TP (Ego)  :", round(mean(acc$p_1_ego,   na.rm = TRUE), 3), "\n")
cat("  TN (Ego)  :", round(mean(acc$p_0_ego,   na.rm = TRUE), 3), "\n")
cat("  TP (Alter):", round(mean(acc$p_1_alter, na.rm = TRUE), 3), "\n")
cat("  TN (Alter):", round(mean(acc$p_0_alter, na.rm = TRUE), 3), "\n")

acc_mat <- as.matrix(acc[, c("p_0_ego", "p_1_ego", "p_0_alter", "p_1_alter")])
boxplot(
  acc_mat,
  names = c("TN (Ego)", "TP (Ego)", "TN (Alter)", "TP (Alter)"),
  ylab = "Probability",
  main = "Individual Accuracy Rates",
  col = c("#3498db", "#2980b9", "#e74c3c", "#c0392b"),
  border = "gray30"
)
```

## Perceptual Structure

The imaginary census and reciprocity error counts summarize which types
of misperception are most prevalent across all perceivers.

```{r structure}
krack_census <- count_imaginary_census(krack_graph)
krack_recip  <- count_recip_errors(krack_graph)

cat("Top imaginary census motifs:\n")
print(head(krack_census, 5))
cat("\nReciprocity errors:\n")
print(head(krack_recip, 5))
```

## Organizational Takeaways

```{r takeaways}
density <- sum(advice_matrix) / (n_people * (n_people - 1))
avg_tp  <- mean(acc$p_1_ego, na.rm = TRUE)
best    <- which.max(acc$p_1_ego)
worst   <- which.min(acc$p_1_ego)

cat("Employees:              ", n_people, "\n")
cat("Advice network density: ", round(density, 3), "\n")
cat("Avg TP (Ego):           ", round(avg_tp, 3), "\n")
cat("Most accurate perceiver:", best,
    "(", round(acc$p_1_ego[best], 3), ")\n")
cat("Least accurate perceiver:", worst,
    "(", round(acc$p_1_ego[worst], 3), ")\n")
```

