---
title: "Real-World Example: Krackhardt Advice Network"
author: "Sima Najafzadehkhoei"
date: today
vignette: >
  %\VignetteIndexEntry{Real-world example of imaginarycss}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# Krackhardt Advice Network

This vignette demonstrates a complete `imaginarycss` workflow using the
classic Krackhardt (1987) dataset of 21 managers in a high-tech company.
We construct a CSS object from the advice network and its individual
perceptions, then compute accuracy rates and the imaginary census.

```{r setup}
library(imaginarycss)
```

## Data and Perceptions

The package ships the advice network as a data frame (`krack_advice`) and
a list of 21 perception matrices (`krack_advice_perceptions`).
We convert the edge-list to an adjacency matrix and bundle it with the
perception data into a `barry_graph` object.

```{r data}
data(krack_advice)
data(krack_advice_perceptions)

# Convert edge-list data frame to adjacency matrix
n_people <- max(c(krack_advice$from, krack_advice$to))
advice_matrix <- matrix(0L, nrow = n_people, ncol = n_people)
advice_matrix[cbind(krack_advice$from, krack_advice$to)] <- krack_advice$value

krack_graph <- new_barry_graph(c(list(advice_matrix), krack_advice_perceptions))
```

## Individual Accuracy

The function `tie_level_accuracy()` decomposes each perceiver's accuracy
into four probabilities, separating ego-involved from alter-only dyads and
true positives from true negatives.

```{r accuracy, fig.height=5, fig.width=6}
acc <- tie_level_accuracy(krack_graph)

# Mean accuracy rates as a data.frame
data.frame(
  Measure = c("TP (Ego)", "TN (Ego)", "TP (Alter)", "TN (Alter)"),
  Mean    = round(c(
    mean(acc$p_1_ego,   na.rm = TRUE),
    mean(acc$p_0_ego,   na.rm = TRUE),
    mean(acc$p_1_alter, na.rm = TRUE),
    mean(acc$p_0_alter, na.rm = TRUE)
  ), 3)
)

acc_mat <- as.matrix(acc[, c("p_0_ego", "p_1_ego", "p_0_alter", "p_1_alter")])
boxplot(
  acc_mat,
  names = c("TN (Ego)", "TP (Ego)", "TN (Alter)", "TP (Alter)"),
  ylab = "Probability",
  main = "Individual Accuracy Rates",
  col = c("#3498db", "#2980b9", "#e74c3c", "#c0392b"),
  border = "gray30"
)
```

## Perceptual Structure

The imaginary census and reciprocity error counts summarize which types
of misperception are most prevalent across all perceivers.

```{r}
#| label: structure
#| collapse: true
krack_census <- count_imaginary_census(krack_graph)
krack_recip  <- count_recip_errors(krack_graph)

# Top imaginary census motifs
head(krack_census, 5)

# Reciprocity errors
head(krack_recip, 5)
```

## Organizational Takeaways

```{r}
#| label: takeaways
#| collapse: true
density <- sum(advice_matrix) / (n_people * (n_people - 1))
avg_tp  <- mean(acc$p_1_ego, na.rm = TRUE)
best    <- which.max(acc$p_1_ego)
worst   <- which.min(acc$p_1_ego)

# Network summary
data.frame(
  Statistic = c("Employees", "Advice network density",
                "Avg TP (Ego)", "Most accurate perceiver",
                "Least accurate perceiver"),
  Value     = c(n_people, round(density, 3), round(avg_tp, 3),
                paste0(best, " (", round(acc$p_1_ego[best], 3), ")"),
                paste0(worst, " (", round(acc$p_1_ego[worst], 3), ")"))
)
```

