---
title: "Introduction to imaginarycss"
author: "George Vega Yon"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    fig-width: 7
    fig-height: 5
    theme: none
    minimal: true
    embed-resources: true
vignette: >
  %\VignetteIndexEntry{Introduction to imaginarycss}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
   echo = TRUE,
   collapse = TRUE,
   comment = "#>",
   fig.path = "man/figures/vignette-",
   out.width = "100%",
   dpi = 300,
   warning = FALSE,
   message = FALSE
 )
```



# Analyzing Perceptual Errors

Understanding how and why people misperceive social networks is at the heart of CSS analysis. The `imaginarycss` package provides several sophisticated measures to quantify different types of perceptual errors.

## Reciprocity Errors

Reciprocity - whether relationships are mutual - is a fundamental property of social networks. When person A says they're friends with person B, is the feeling mutual? Reciprocity errors occur when individuals misperceive the reciprocal nature of relationships.

### Types of Reciprocity Errors

1. **Assumed Reciprocity**: Thinking a one-way relationship is mutual
2. **Missed Reciprocity**: Thinking a mutual relationship is one-way  
3. **False Asymmetry**: Seeing asymmetry where there's actually reciprocity
4. **False Symmetry**: Seeing reciprocity where there's actually asymmetry

### Why Reciprocity Matters

- **Social Expectations**: We expect important relationships to be reciprocal
- **Cognitive Load**: It's easier to remember "A and B are friends" than "A likes B but B doesn't like A"
- **Projection**: We assume others feel the same way about us as we do about them

```{r reciprocity-errors}
# Compute reciprocity errors
recip_errors <- count_recip_errors(graph)
print(recip_errors)

# Visualize the distribution
ggplot(recip_errors, aes(x = reorder(name, value), y = value)) +
  geom_col(
    fill = css_colors["primary"], 
    alpha = 0.8,
    width = 0.7
  ) +
  geom_text(
    aes(label = value), 
    hjust = -0.2, 
    color = css_colors["dark"],
    size = 3.5,
    fontface = "bold"
  ) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "🔄 Reciprocity Errors by Type",
    subtitle = "Understanding how people misperceive mutual relationships",
    x = "Error Type",
    y = "Count",
    caption = "Higher values indicate more frequent misperceptions"
  ) +
  theme(
    axis.text.y = element_text(size = 10),
    plot.title = element_text(color = css_colors["primary"])
  )
```

## Imaginary Census

The imaginary census provides a comprehensive classification of all possible perceptual errors. Think of it as a "taxonomy of mistakes" - every possible way someone can misperceive a dyadic relationship gets its own category.

### The 10 Categories Explained

The census classifies every dyad (pair of people) based on:
1. **What actually exists** (null, asymmetric, or reciprocal tie)
2. **What the perceiver thinks exists** (null, asymmetric, or reciprocal tie)

This creates a 3×3 matrix of possibilities, but some combinations are subdivided:

#### **Null Dyads** (no actual connection):
- **(01) Accurate Null**: Correctly perceives no connection
- **(02) Partial False Positive**: Sees one-way connection where none exists
- **(03) Complete False Positive**: Sees mutual connection where none exists

#### **Asymmetric Dyads** (one-way connection):
- **(04) Partial False Negative**: Misses the connection entirely  
- **(05) Accurate Asymmetric**: Correctly sees one-way connection
- **(06) Mixed Asymmetric**: Sees connection but gets direction wrong
- **(07) Partial False Positive**: Sees mutual connection where only one-way exists

#### **Reciprocal Dyads** (mutual connection):
- **(08) Complete False Negative**: Misses the connection entirely
- **(09) Partial False Negative**: Sees only one direction of mutual tie
- **(10) Accurate Reciprocal**: Correctly perceives mutual connection

### Cognitive Implications

Different error patterns suggest different underlying cognitive processes:
- **High false positives**: "Rose-colored glasses" - seeing connections everywhere
- **High false negatives**: "Tunnel vision" - missing connections that exist  
- **Direction errors**: Difficulty with asymmetric relationships
- **Reciprocity bias**: Tendency to assume relationships are mutual

```{r imaginary-census}
# Compute full imaginary census
census <- count_imaginary_census(graph)
print(census)

# Clean up names for better visualization
# Convert to regular data frame to avoid class conflicts
census_df <- census[, names(census)]  # Strip custom class
class(census_df) <- "data.frame"

census_df$name_clean <- gsub("\\([0-9]+\\)$", "", census_df$name)
census_summary <- census_df %>%
  filter(grepl("^\\([0-9]", name)) %>%
  mutate(
    error_type = case_when(
      grepl("Accurate null", name_clean) ~ "Accurate Null",
      grepl("Partial false positive \\(null\\)", name_clean) ~ "Partial FP (Null)",
      grepl("Complete false positive \\(null\\)", name_clean) ~ "Complete FP (Null)",
      grepl("Partial false negative \\(assym\\)", name_clean) ~ "Partial FN (Asymm)",
      grepl("Accurate assym", name_clean) ~ "Accurate Asymm",
      grepl("Mixed assym", name_clean) ~ "Mixed Asymm",
      grepl("Partial false positive \\(assym\\)", name_clean) ~ "Partial FP (Asymm)",
      grepl("Complete false negative \\(full\\)", name_clean) ~ "Complete FN (Full)",
      grepl("Partial false negative \\(full\\)", name_clean) ~ "Partial FN (Full)",
      grepl("Accurate full", name_clean) ~ "Accurate Full",
      TRUE ~ "Other"
    )
  )

# Create a beautiful color palette for error types
error_colors <- c(
  "Accurate Null" = "#27ae60",           # Green for accurate
  "Accurate Asymm" = "#2ecc71",          # Light green
  "Accurate Full" = "#16a085",           # Teal
  "Partial FP (Null)" = "#f39c12",       # Orange for mild errors
  "Partial FN (Asymm)" = "#e67e22",      # Dark orange
  "Partial FN (Full)" = "#d35400",       # Red-orange
  "Complete FP (Null)" = "#e74c3c",      # Red for serious errors
  "Complete FN (Full)" = "#c0392b",      # Dark red
  "Partial FP (Asymm)" = "#8e44ad",      # Purple
  "Mixed Asymm" = "#9b59b6",             # Light purple
  "Other" = "#95a5a6"                    # Gray for other
)

# Visualize error distribution
ggplot(census_summary, aes(x = reorder(error_type, value), y = value)) +
  geom_col(aes(fill = error_type), alpha = 0.9, width = 0.8) +
  geom_text(
    aes(label = ifelse(value > 0, value, "")), 
    hjust = -0.1, 
    color = "#2c3e50",
    size = 3.2,
    fontface = "bold"
  ) +
  scale_fill_manual(values = error_colors) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "🎯 Distribution of Perceptual Errors",
    subtitle = "Comprehensive taxonomy of social perception mistakes",
    x = "Error Type",
    y = "Count",
    caption = "Green = Accurate perceptions, Orange/Red = False positives/negatives, Purple = Mixed errors"
  ) +
  guides(fill = "none") +
  theme(
    axis.text.y = element_text(size = 9),
    plot.title = element_text(color = css_colors["accent"])
  )
```

# Analyzing Individual Accuracy

Individual differences in social perception are often more interesting than aggregate patterns. Some people are "social radar" - exceptionally good at reading social networks. Others are "social blind spots" - systematically missing important patterns.

## Tie-Level Accuracy

This analysis breaks down accuracy into four key measures, distinguishing between different types of relationships and different types of correct judgments.

### The Four Accuracy Measures

1. **True Positive (Ego)**: Correctly identifying ties that involve the perceiver
   - "I correctly know who I'm connected to"
   - High scores suggest good self-awareness

2. **True Negative (Ego)**: Correctly identifying non-ties that involve the perceiver  
   - "I correctly know who I'm NOT connected to"
   - High scores suggest realistic self-assessment

3. **True Positive (Alter)**: Correctly identifying ties between other people
   - "I correctly know who else is connected"
   - High scores suggest good social observation skills

4. **True Negative (Alter)**: Correctly identifying non-ties between other people
   - "I correctly know who else is NOT connected"  
   - High scores suggest avoiding false gossip/assumptions

### Why This Decomposition Matters

**Ego vs. Alter Accuracy:**
- People are typically more accurate about their own relationships (ego) than others' relationships (alter)
- But some individuals are better "social observers" and excel at alter accuracy
- The ratio can reveal whether someone is self-focused vs. socially aware

**True Positive vs. True Negative Accuracy:**
- True positive accuracy: "Am I good at spotting connections?"
- True negative accuracy: "Am I good at recognizing when there's no connection?"
- Different cognitive processes may underlie these two abilities

Understanding individual differences in perception accuracy is crucial:

```{r tie-accuracy}
# Compute tie-level accuracy for each individual
accuracy <- tie_level_accuracy(graph)
print(accuracy)

# Reshape data for visualization
# Convert to regular data frame first to avoid class conflicts
accuracy_df <- accuracy[, names(accuracy)]  # Strip custom class
class(accuracy_df) <- "data.frame"

accuracy_long <- accuracy_df %>%
  select(-k) %>%
  tidyr::pivot_longer(
    everything(),
    names_to = "measure",
    values_to = "probability"
  ) %>%
  mutate(
    measure_clean = case_when(
      measure == "p_0_ego" ~ "True Negative (Ego)",
      measure == "p_1_ego" ~ "True Positive (Ego)", 
      measure == "p_0_alter" ~ "True Negative (Alter)",
      measure == "p_1_alter" ~ "True Positive (Alter)"
    )
  )

# Create color palette for accuracy measures
accuracy_colors <- c(
  "True Negative (Ego)" = "#3498db",     # Blue
  "True Positive (Ego)" = "#2980b9",     # Dark blue  
  "True Negative (Alter)" = "#e74c3c",   # Red
  "True Positive (Alter)" = "#c0392b"    # Dark red
)

# Visualize accuracy measures
p1 <- ggplot(accuracy_long, aes(x = measure_clean, y = probability)) +
  geom_violin(
    aes(fill = measure_clean),
    alpha = 0.7,
    trim = FALSE,
    scale = "width"
  ) +
  geom_boxplot(
    width = 0.2,
    alpha = 0.8,
    outlier.alpha = 0.6,
    outlier.size = 1.5
  ) +
  geom_jitter(
    width = 0.15, 
    alpha = 0.6, 
    size = 2,
    color = "#2c3e50"
  ) +
  scale_fill_manual(values = accuracy_colors) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "🎯 Individual-Level Accuracy Rates",
    subtitle = "How well do people perceive different types of relationships?",
    x = "Accuracy Measure",
    y = "Accuracy Rate (%)",
    caption = "Blue = Ego relationships (involving self), Red = Alter relationships (between others)"
  ) +
  guides(fill = "none") +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 10),
    plot.title = element_text(color = css_colors["info"]),
    panel.grid.major.x = element_blank()
  )

print(p1)
```

## References

- Krackhardt, D. (1987). Cognitive social structures. *Social Networks*, 9(2), 109-134.
- Vega Yon, G. G., & Demarais, A. (2022). Exponential random graph models for little networks. *Social Networks*, 70, 181-195.

---

*This vignette was created with `imaginarycss` version `r packageVersion("imaginarycss")`.*