---
title: "Real-world example of imaginarycss"
author: "Sima Najafzadehkhoei"
date: today
format: html
vignette: >
  %\VignetteIndexEntry{Real-world example of imaginarycss}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---


# Real-World Example: Krackhardt Network

Let's demonstrate the package with a realistic example using the actual Krackhardt organizational network data. The Krackhardt advice network is a classic dataset in organizational behavior research.

## Understanding Organizational Networks

**The Setting**: A high-tech company where researchers measured both:
- **Actual advice relationships**: Who actually seeks advice from whom (observed)
- **Perceived advice relationships**: Who each employee thinks seeks advice from whom

**Why This Matters**:
- **Information flow**: Advice networks determine how knowledge spreads
- **Influence patterns**: Perceived networks affect who people try to influence  
- **Organizational design**: Managers need accurate views of informal structure
- **Change management**: Interventions should target actual vs. perceived bottlenecks

## The Cognitive Challenge

Organizational advice networks present unique perceptual challenges:

1. **Visibility**: Advice-seeking often happens privately (email, closed-door meetings)
2. **Asymmetry**: Advice relationships are inherently directional and often non-reciprocal
3. **Status effects**: People may misperceive advice patterns based on formal hierarchy
4. **Wishful thinking**: Employees might report advice patterns they wish existed

```{r krackhardt-setup, eval=FALSE}
# Load the actual Krackhardt data
library(imaginarycss)
data(krack_advice)
data(krack_attributes)

# Convert the adjacency matrix format to actual adjacency matrix
library(reshape2)
advice_matrix <- acast(krack_advice, from ~ to, value.var = "value", fill = 0)

# For this example, we'll use the advice network as the "true" network
# and create some realistic perception errors based on organizational factors
n_people <- nrow(advice_matrix)

# Create perception errors based on organizational factors
create_realistic_perceptions <- function(true_net, attributes) {
  n <- nrow(true_net)
  perceived_networks <- list()
  
  for (i in 1:n) {
    perceived <- true_net
    
    # Different types of systematic errors:
    # 1. Status bias: overestimate advice seeking to high-status people
    # 2. Visibility bias: miss private advice relationships
    # 3. Reciprocity bias: assume advice relationships are reciprocal
    
    for (j in 1:n) {
      for (k in 1:n) {
        if (i != j && i != k) {  # Person i perceiving j->k relationship
          
          # Visibility bias: miss some true relationships (false negatives)
          if (true_net[j, k] == 1 && runif(1) < 0.2) {
            perceived[j, k] <- 0
          }
          
          # Status bias: add false relationships to high-degree nodes
          if (true_net[j, k] == 0 && sum(true_net[, k]) > median(colSums(true_net)) && runif(1) < 0.15) {
            perceived[j, k] <- 1
          }
          
          # Reciprocity bias: if j seeks advice from k, assume k seeks from j
          if (true_net[j, k] == 1 && true_net[k, j] == 0 && runif(1) < 0.1) {
            perceived[k, j] <- 1
          }
        }
      }
    }
    
    # Ensure no self-loops
    diag(perceived) <- 0
    perceived_networks[[i]] <- perceived
  }
  
  return(perceived_networks)
}

# Create realistic perceptions
perceived_networks <- create_realistic_perceptions(advice_matrix, krack_attributes)

# Combine true network with all individual perceptions
all_networks <- c(list(advice_matrix), perceived_networks)
krack_graph <- new_barry_graph(all_networks)

# Add some network metadata
attr(krack_graph, "network_type") <- "advice"
attr(krack_graph, "organization") <- "high_tech_company"
attr(krack_graph, "n_employees") <- n_people
```

```{r krackhardt-analysis, eval=FALSE}
# Analyze the organizational network
krack_accuracy <- tie_level_accuracy(krack_graph)

# Summary statistics
accuracy_df <- krack_accuracy[, names(krack_accuracy)]
class(accuracy_df) <- "data.frame"

summary_stats <- accuracy_df %>%
  select(-k) %>%
  summarise(
    across(everything(), list(mean = mean, sd = sd), na.rm = TRUE)
  )

print("Overall Accuracy Statistics:")
print(summary_stats)

# Individual differences in accuracy with organizational context
individual_plot <- ggplot(krack_accuracy, aes(x = factor(k))) +
  geom_point(
    aes(y = p_1_ego, color = "True Positive (Ego)"), 
    size = 3.5,
    alpha = 0.8
  ) +
  geom_point(
    aes(y = p_0_ego, color = "True Negative (Ego)"), 
    size = 3.5,
    alpha = 0.8
  ) +
  geom_point(
    aes(y = p_1_alter, color = "True Positive (Alter)"), 
    size = 3.5,
    alpha = 0.8
  ) +
  geom_point(
    aes(y = p_0_alter, color = "True Negative (Alter)"), 
    size = 3.5,
    alpha = 0.8
  ) +
  geom_smooth(
    aes(y = p_1_ego, color = "True Positive (Ego)"),
    method = "loess",
    se = FALSE,
    size = 1,
    alpha = 0.6
  ) +
  geom_smooth(
    aes(y = p_0_ego, color = "True Negative (Ego)"),
    method = "loess", 
    se = FALSE,
    size = 1,
    alpha = 0.6
  ) +
  scale_color_manual(
    values = c(
      "True Positive (Ego)" = "#2980b9",
      "True Negative (Ego)" = "#3498db", 
      "True Positive (Alter)" = "#c0392b",
      "True Negative (Alter)" = "#e74c3c"
    ),
    name = "Accuracy Measure"
  ) +
  scale_y_continuous(
    labels = scales::percent_format(),
    limits = c(0, 1)
  ) +
  labs(
    title = "🧠 Individual Accuracy in High-Tech Company",
    subtitle = "How well each employee perceives the advice network",
    x = "Employee ID",
    y = "Accuracy Rate (%)",
    caption = "Data: Krackhardt's high-tech managers study"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.box.background = element_rect(color = "#bdc3c7", fill = "#ffffff", alpha = 0.8)
  ) +
  guides(
    color = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      nrow = 2,
      override.aes = list(size = 4)
    )
  )

print(individual_plot)

# Network-level analysis using real data
krack_census <- count_imaginary_census(krack_graph)
krack_recip <- count_recip_errors(krack_graph)

print("Imaginary Structure Census:")
print(krack_census)

print("Reciprocity Errors:")
print(krack_recip)

# Compare with other network types (if available)
# You could extend this to use friendship and reporting networks too
if (exists("krack_friendship") && exists("krack_reports")) {
  
  # Load and process friendship network
  friendship_matrix <- acast(krack_friendship, from ~ to, value.var = "value", fill = 0)
  reports_matrix <- acast(krack_reports, from ~ to, value.var = "value", fill = 0)
  
  # Compare network densities
  network_comparison <- data.frame(
    Network = c("Advice", "Friendship", "Reports"),
    Density = c(
      sum(advice_matrix) / (nrow(advice_matrix) * (nrow(advice_matrix) - 1)),
      sum(friendship_matrix) / (nrow(friendship_matrix) * (nrow(friendship_matrix) - 1)),
      sum(reports_matrix) / (nrow(reports_matrix) * (nrow(reports_matrix) - 1))
    ),
    Reciprocity = c(
      sum(advice_matrix * t(advice_matrix)) / sum(advice_matrix),
      sum(friendship_matrix * t(friendship_matrix)) / sum(friendship_matrix),
      sum(reports_matrix * t(reports_matrix)) / sum(reports_matrix)
    )
  )
  
  print("Network Type Comparison:")
  print(network_comparison)
}

# Performance benchmarking with real data
print("Performance with real organizational data:")
system.time({
  sample_css_network(krack_graph)
})

# Practical insights for the organization
cat("\n=== ORGANIZATIONAL INSIGHTS ===\n")
cat("Network Size:", n_people, "employees\n")
cat("Advice Density:", round(sum(advice_matrix) / (n_people * (n_people - 1)), 3), "\n")
cat("Average Perception Accuracy:", round(mean(krack_accuracy$p_1_ego, na.rm = TRUE), 3), "\n")

# Identify most/least accurate perceivers
most_accurate <- which.max(krack_accuracy$p_1_ego)
least_accurate <- which.min(krack_accuracy$p_1_ego)

cat("Most accurate perceiver: Employee", most_accurate, 
    "(", round(krack_accuracy$p_1_ego[most_accurate], 3), "accuracy )\n")
cat("Least accurate perceiver: Employee", least_accurate, 
    "(", round(krack_accuracy$p_1_ego[least_accurate], 3), "accuracy )\n")
```

## Organizational Applications

This analysis reveals several practical insights:

1. **Individual Differences**: Some employees are much better at reading the advice network than others
2. **Systematic Biases**: People tend to overestimate advice-seeking to high-status individuals
3. **Network Visibility**: Private advice relationships are often missed in perceptions
4. **Intervention Targets**: Focus training on employees with lowest accuracy scores
## References

- Krackhardt, D. (1987). Cognitive social structures. *Social Networks*, 9(2), 109-134.
- Vega Yon, G. G., & Demarais, A. (2022). Exponential random graph models for little networks. *Social Networks*, 70, 181-195.

---

*This vignette was created with `imaginarycss` version `r packageVersion("imaginarycss")`.*

