---
title: "Introduction to imaginarycss"
author: "George Vega Yon"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    fig-width: 7
    fig-height: 5
    theme: none
    minimal: true
    embed-resources: true
vignette: >
  %\VignetteIndexEntry{Introduction to imaginarycss}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
   echo = TRUE,
   collapse = TRUE,
   comment = "#>",
   fig.path = "man/figures/vignette-",
   out.width = "100%",
   dpi = 300,
   warning = FALSE,
   message = FALSE
 )
```
# Generating Null Models

Statistical testing in network analysis requires careful null models. We can't just use random networks because individual perceptual biases aren't random - they're systematic and person-specific.

## The Challenge of Network Null Models

Traditional approaches might use:
- **Random networks**: But real perception isn't random
- **Erdős-Rényi models**: But they ignore individual differences
- **Configuration models**: But they don't capture perceptual biases

### Our Solution: Accuracy-Based Sampling

The `imaginarycss` approach uses individual accuracy rates to generate realistic null distributions:

1. **Measure each person's accuracy** for different types of relationships
2. **Use these rates as probabilities** for generating simulated perceptions
3. **Preserve individual differences** while randomizing specific errors
4. **Generate many samples** to build null distributions

This approach asks: "If someone has X% accuracy rate for ego ties and Y% accuracy for alter ties, what range of error patterns would we expect by chance?"

## Sampling Perceived Networks

The sampling process works as follows:

1. **Start with the true network** as baseline
2. **For each perceiver and each potential tie:**
   - If it's an ego tie: Use their ego accuracy rate
   - If it's an alter tie: Use their alter accuracy rate  
   - If it should exist: Use true positive rate as probability
   - If it shouldn't exist: Use true negative rate as probability
3. **Generate binary outcome** based on these probabilities
4. **Repeat many times** to create distribution

```{r sampling}
# Sample a single network based on accuracy rates
set.seed(123)
sampled_networks <- sample_css_network(graph, keep_baseline = TRUE)

cat("Number of sampled networks:", length(sampled_networks), "\n")
cat("Dimensions of each network:", dim(sampled_networks[[1]]), "\n")

# Compare original vs sampled network density
original_density <- sum(sampled_networks[[1]]) / (4 * 3)  # exclude diagonal
sampled_density <- sum(sampled_networks[[2]]) / (4 * 3)

cat("Original network density:", round(original_density, 3), "\n")
cat("Sampled network density:", round(sampled_density, 3), "\n")
```

## Generating Multiple Samples for Statistical Testing

The power of null models comes from generating many samples to understand the distribution of outcomes we'd expect by chance alone.

### The Statistical Logic

1. **Observed Pattern**: We see X false positives in our data
2. **Null Hypothesis**: This is just what we'd expect given individual accuracy rates
3. **Null Distribution**: Generate 1000+ samples using accuracy rates as probabilities
4. **P-value**: What proportion of null samples have ≥ X false positives?
5. **Interpretation**: If p < 0.05, the observed pattern is unlikely due to chance

### Why This Approach Works

- **Preserves individual differences**: Each person's bias level stays realistic
- **Controls for network density**: Null models have similar overall connection rates
- **Accounts for position effects**: Ego vs. alter accuracy differences are maintained
- **Enables fair comparison**: Only the specific pattern of errors is randomized

```{r multiple-samples}
# Generate multiple samples for null distribution
set.seed(456)
n_samples <- 100

# Note: For this example, we'll use a simpler approach
# since the full null sampling may require additional setup
null_samples <- replicate(n_samples, {
  sample_css_network(graph, keep_baseline = FALSE)
}, simplify = FALSE)

# For demonstration, let's compute a simple metric instead of full census
# We'll count the total number of ties in each sampled network
null_densities <- sapply(null_samples, function(nets) {
  if (length(nets) > 0) {
    sum(nets[[1]]) / (4 * 3)  # Network density
  } else {
    0
  }
})

# Compare with observed density
observed_density <- sum(sampled_networks[[2]]) / (4 * 3)

# Calculate p-value first
p_value <- mean(null_densities >= observed_density)
cat("P-value for network density:", round(p_value, 4), "\n")

# Visualization
null_df <- data.frame(
  density = null_densities,
  type = "Null Distribution"
)

# Create beautiful density comparison plot
density_plot <- ggplot(null_df, aes(x = density)) +
  geom_histogram(
    bins = 25, 
    alpha = 0.8, 
    fill = css_colors["primary"], 
    color = "white",
    size = 0.3
  ) +
  geom_vline(
    xintercept = observed_density, 
    color = css_colors["secondary"], 
    linetype = "dashed", 
    size = 1.2
  ) +
  geom_density(
    aes(y = after_stat(density) * length(null_densities) * 0.05),
    color = css_colors["accent"],
    size = 1.5,
    alpha = 0.7
  ) +
  annotate(
    "label", 
    x = observed_density + 0.02, 
    y = Inf, 
    label = paste("Observed\n", round(observed_density, 3)),
    vjust = 1.2,
    hjust = 0,
    color = css_colors["secondary"],
    fill = "white",
    size = 3.5,
    fontface = "bold",
    label.padding = unit(0.3, "lines")
  ) +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "📊 Null Distribution of Network Density",
    subtitle = "Comparing observed patterns against chance expectations",
    x = "Network Density (%)",
    y = "Frequency",
    caption = paste("P-value:", round(p_value, 4), "| Based on", n_samples, "null samples")
  ) +
  theme(
    plot.title = element_text(color = css_colors["primary"]),
    panel.grid.major.x = element_line(color = "#ecf0f1", size = 0.5),
    panel.grid.major.y = element_line(color = "#ecf0f1", size = 0.5)
  )

print(density_plot)
```

## References

- Krackhardt, D. (1987). Cognitive social structures. *Social Networks*, 9(2), 109-134.
- Vega Yon, G. G., & Demarais, A. (2022). Exponential random graph models for little networks. *Social Networks*, 70, 181-195.

---

*This vignette was created with `imaginarycss` version `r packageVersion("imaginarycss")`.*